---
title: "M4: Outliers and Missing Data - Wind Speed Data"
author: "Chloe Young"
output: pdf_document
---


## Setting R code chunk options

First R code chunk is used for setting the options for all R code chunks. The choice echo=TRUE means both code and output will appear on report, include = FALSE neither code nor output is printed.

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

Second R code chunk is for loading packages. By setting message = FALSE, the code will appear but not the output. 

```{r package, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(Kendall)
library(tseries)
library(zoo)

#New packages for M4
#install.packages("outliers")
library(outliers)
#install.packages("tidyverse")
library(tidyverse)

```

## Importing and preparing data

Today we will work with wind speed data. 

```{r}

#Importing time series data from text file#
wind_data <- read.csv(file="/home/guest/TSA_Sp25/Data/Wind_Speed_PortArthurTX_Processed.csv",header=TRUE,stringsAsFactors = TRUE)

#creating data object
wind_data$DATE <- ym(wind_data$DATE)

```


## Transforming data into time series object

Many of the functions we will use require a time series object. You can transform your data in a time series using the function *ts()*. 

```{r}
ts_wind_data <- ts(wind_data$AWND.1,frequency=12, start=c(1984,1)) 

autoplot(ts_wind_data) +
  ylab("Average Wind Speed") 

summary(wind_data$AWND.1) #tells you how many NA's there are
```


## Using pre-built function to clean series

tsclean() from package 'forecast' identify and replace outliers and missing values in a time series using linear interpolation. This should be used with some caution!!

```{r}

clean_wind_data <- tsclean(ts_wind_data)  #object should be a time series

autoplot(clean_wind_data, series="Wind Clean") +
  autolayer(ts_wind_data, series="Wind Raw") +
  ylab("Average Wind Speed") 

#Note we only replaced missing data, algorithm did not identify any outliers for this data set.

#Take caution because we know we want to replace missing values, not outliers. 
  

```

If you want to handle missing data, but do not want to remove outliers, here are two other opion that will replace them using linear interpolation.

`forecast::na.interp()`
`zoo::na.approx()`

```{r}
#Using na.interp
clean_wind_data2 <- forecast::na.interp(ts_wind_data)  #object should be a time series

autoplot(clean_wind_data2, series="Wind Clean") +
  autolayer(ts_wind_data, series="Wind Raw") +
  ylab("Average Wind Speed") 

```

```{r}

#Using na.approx
clean_wind_data3 <- zoo::na.approx(ts_wind_data)  #object should be a time series

autoplot(clean_wind_data3, series="Wind Clean") +
  autolayer(ts_wind_data, series="Wind Raw") +
  ylab("Average Wind Speed") 

```

The results from na.interp and na.approx only have 4 spots, which makes sense because we identified that there were only 4 NAs. This confirms that the clean function also removed outliers since there are more red spots. 

